Identifying Microlensing Events in Large, Non-Uniformly Sampled Surveys: The Case of the Palomar Transient Factory

### 2012-08-14 text/paper/ptf_microlensing_paper.tex
move paper into paper directory, add figures directory

\title{Identifying Microlenses in Large, Non-Uniformly Sampled Surveys: The Case
 of the Palomar Transient Factory}
\author{Adrian~M.~Price-Whelan\altaffilmark{\ref{col}}, Marcel~A.~Ag\"ueros\altaffilmark{\ref{col}}, Amanda Fournier\altaffilmark{\ref{ucsb}}, Rachel Street\altaffilmark{\ref{lcogt}}, Eran Ofek\altaffilmark{\ref{weiz},\ref{eins}}, David Levitan\altaffilmark{\ref{calt}}, Joshua S.\ Bloom\altaffilmark{\ref{cal}}, S.\ Bradley Cenko\altaffilmark{\ref{cal}}, Mansi M.\ Kasliwal\altaffilmark{\ref{calt}}, Shrinivas R.\ Kulkarni\altaffilmark{\ref{calt}}, Nicholas~M.~Law\altaffilmark{\ref{to},\ref{dun}}, Peter Nugent\altaffilmark{\ref{lbnl}}, Dovi Poznanski\altaffilmark{\ref{cal},\ref{calt},\ref{eins}}, Robert M.\ Quimby\altaffilmark{\ref{calt}}}

@@

The PTF camera is the 12K$\times$8K mosaic camera formerly at the Canada-France-Hawaii Telescope. The camera has 11 working chips, 10$^7$ pixels, and a 7.26 deg$^2$ field-of-view \citep{rahmer2008}. Under median seeing conditions (1.1$\arcsec$), observations in Mould $R$ or Sloan Digital Sky Survey (SDSS) $g$ achieve 2.0$\arcsec$ full-width half-maximum images and reach 5$\sigma$ magnitudes of $R \approx 21.0$ and $g \approx 21.3$ mag in a standard 60~s exposure \citep{nick2010}. As of August 9, 2012, the PTF footprint included $\apwsim$10,100 (2600) deg$^2$ imaged $>$25 ($>$100) times in $R$ and $\apwsim$2200 (170) deg$^2$ imaged that often in $g$.

[UPDATE NUMBERS FOR FOOTPRINT, INCLUDE FIGURE?]

@@

The average number of light curves with $>$25 ``good'' observations is $\apwsim 50,000$ per PTF field, with $\apwsim64,000,000$  ``good'' light curves in total. By our estimates, utilizing all cores on an 8-core 2.0 GHz compute, it would take [TODO: estimate this number] to fit even the simplest 5-parameter microlensing event model to each of these light curves using \inlinecode{ANSI-C} wrapped in \inlinecode{Python}. Previous studies have had the advantage of pre-selecting candidates via difference image analysis, but have utilized a variety of selection techniques for flagging candidate microlensing light curves after this step. However, most of these methods still rely on fitting a point-source, point-lens microlensing model to each light curve and performing some selection based on the $\chi^2$ value of this fit (cite Sumi et al. 2006, MACHO work, ?). More recent work has begun to apply machine learning algorithms to the general light curve classification problem (cite Joey Richards), however this is still a work in progress. We set out to find a fast, robust selection method that [TODO:].

@@

- Discuss $\Delta \chi^2$ method, downfalls (speed, not robust -- how do you define the threshold?)

@@

%\bibliographystyle{apj}
%\bibliography{apj-jour,refs}
%
%\begin{figure}
%	\centering
%	\caption{Computed detection efficiency as a function of number of observations for a fixed baseline of 365 days. This plot uses entirely simulated light curves.}
%    \includegraphics[width=1.0\textwidth]{figures/detection_efficiency_figure.pdf}
%    \label{fig:detection_eff}
%\end{figure}	
%
%\begin{figure}
%	\centering
%	\caption{A visualization of the different sampling regimes of six different PTF fields over the same one year baseline. Darker points mean more data (more exposures). Some fields have roughly uniform coverage, where others are very dense during some weeks and not others. }
%    \includegraphics[width=1.0\textwidth]{figures/ptf_sampling_figure.pdf}
%    \label{fig:light_curves}
%\end{figure}
%
%\begin{figure}
%	\centering
%	\caption{Example of how the variability indices may be used to separate microlensing events from periodic or flat light curves.}
%    \includegraphics[width=1.0\textwidth]{figures/J_vs_K_Con_figure.png}
%    \label{fig:var_idx1}
%\end{figure}
%
%\begin{figure}[htbp]
%	\centering
%	\caption{Two neighboring light curves. The time scale is the same on both plots. Note the strong similarities in the overall shapes of the data.} 
%	 \includegraphics[width=1.0\textwidth]{figures/bad_data_figure.pdf}
%	\label{fig:lc_correlated}
%\end{figure}

>>>>

\title{Identifying Microlensing Events in Large, Non-Uniformly Sampled Surveys: The Case
 of the Palomar Transient Factory}
\author{Adrian~M.~Price-Whelan\altaffilmark{\ref{col}}, Marcel~A.~Ag\"ueros\altaffilmark{\ref{col}}, Amanda Fournier\altaffilmark{\ref{ucsb}}, Rachel Street\altaffilmark{\ref{lcogt}}, Eran Ofek\altaffilmark{\ref{weiz},\ref{eins}}, David Levitan\altaffilmark{\ref{calt}}, Joshua S.\ Bloom\altaffilmark{\ref{cal}}, S.\ Bradley Cenko\altaffilmark{\ref{cal}}, Mansi M.\ Kasliwal\altaffilmark{\ref{calt}}, Shrinivas R.\ Kulkarni\altaffilmark{\ref{calt}}, Nicholas~M.~Law\altaffilmark{\ref{to},\ref{dun}}, Peter Nugent\altaffilmark{\ref{lbnl}}, Dovi Poznanski\altaffilmark{\ref{cal},\ref{calt},\ref{eins}}, Robert M.\ Quimby\altaffilmark{\ref{calt}}}

@@

The PTF camera is the 12K$\times$8K mosaic camera formerly at the Canada-France-Hawaii Telescope. The camera has 11 working chips, 10$^7$ pixels, and a 7.26 deg$^2$ field-of-view \citep{rahmer2008}. Under median seeing conditions (1.1$\arcsec$), observations in Mould $R$ or Sloan Digital Sky Survey (SDSS) $g$ achieve 2.0$\arcsec$ full-width half-maximum images and reach 5$\sigma$ magnitudes of $R \approx 21.0$ and $g \approx 21.3$ mag in a standard 60~s exposure \citep{nick2010}. As of August 9, 2012, the PTF footprint included $\apwsim$10,100 (2600) deg$^2$ imaged $>$25 ($>$100) times in $R$ and $\apwsim$2200 (170) deg$^2$ imaged that often in $g$ (\figurename~\ref{fig:survey_footprint}).

@@

The average number of light curves per PTF field with $>$25 ``good'' observations is $\apwsim 50,000$, with $\apwsim64,000,000$  ``good'' light curves in total. By our estimates, utilizing all cores on an 8-core 2.0 GHz compute, it would take [TODO: estimate this number] to fit even the simplest 5-parameter microlensing event model to each of these light curves using \inlinecode{ANSI-C} wrapped in \inlinecode{Python}. Previous studies have had the advantage of pre-selecting candidates via difference image analysis, automatically limiting their sample to $\apwsim$thousands of light curves. Past work has also utilized a variety of selection techniques for flagging candidate microlensing light curves after these steps, however most of these methods still rely on fitting a point-source, point-lens microlensing model to each light curve and performing some selection based on the $\chi^2$ value of this fit (cite Sumi et al. 2006, MACHO work, ?). More recent work has begun to apply machine learning algorithms to the general light curve classification problem (cite Joey Richards), however this is still a work in progress. We set out to find a fast, robust selection method for identifying microlensing events in big data archives by using a set of variability statistics previously applied to finding variable stars in general (Shin M.-S. et al. ).

[Describe variability indices?]

@@

- Discuss $\Delta \chi^2$ method, downfalls (speed, not robust -- how do you define the threshold?). Show a plot of distribution of $\Delta \chi^2$ with and without simulated events, e.g. the distribution doesn't separate well, then show $\eta$ and how the peak of the distribution moves.

@@

%\bibliographystyle{apj}
%\bibliography{apj-jour,refs}
%
\begin{figure}
	\centering
	\caption{Computed detection efficiency as a function of number of observations for a fixed baseline of 365 days. This plot uses entirely simulated light curves.}
    \includegraphics[width=1.0\textwidth]{figures/detection_efficiency_figure.pdf}
    \label{fig:survey_footprint}
\end{figure}	
%
%\begin{figure}
%	\centering
%	\caption{A visualization of the different sampling regimes of six different PTF fields over the same one year baseline. Darker points mean more data (more exposures). Some fields have roughly uniform coverage, where others are very dense during some weeks and not others. }
%    \includegraphics[width=1.0\textwidth]{figures/ptf_sampling_figure.pdf}
%    \label{fig:light_curves}
%\end{figure}
%
%\begin{figure}
%	\centering
%	\caption{Example of how the variability indices may be used to separate microlensing events from periodic or flat light curves.}
%    \includegraphics[width=1.0\textwidth]{figures/J_vs_K_Con_figure.png}
%    \label{fig:var_idx1}
%\end{figure}
%
%\begin{figure}[htbp]
%	\centering
%	\caption{Two neighboring light curves. The time scale is the same on both plots. Note the strong similarities in the overall shapes of the data.} 
%	 \includegraphics[width=1.0\textwidth]{figures/bad_data_figure.pdf}
%	\label{fig:lc_correlated}
%\end{figure}
